---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

devtools::load_all()
```

# MSMT (ALPHA VERSION)

<!-- badges: start -->
<!-- badges: end -->

The goal of **MSMT** is to facilitate the access to and use of administrative data collected by Czech Ministry of Education, Youth and Sports (further on referred to as *MSMT data*).

## Installation

You can install the working version of **MSMT** from GitHub with:

```{r,eval=FALSE}
library(devtools)
install_github("jmunich/MSMT")
```

To load **MSMT**, run:

```{r, eval=FALSE}
library(MSMT)
```

## **MSMT** data structure

The data is collected on a regular basis using standardized forms filled in by directors of schools and other educational institutions.

There are different types of forms for different types of establishments, each with a unique code. Function **msmt_forms** was designed to facilitate the browsing of available forms and variables. First, we can look at what forms there are:

```{r}
all_forms <-  msmt_forms(language = "en")
all_forms
```

Package **MSMT** currently provides the original labels in Czech and an English translation created using R package **deeplr**.

Each form consists of tables in which the respondents report mostly count data. These tables then represent sets of variables, grouped by a similar topic. We can have a look at the approximation of the original tables, for instance primary school form **M03**, table with Special Educational Needs (SEN, or SVP in the original - as can be seen, the automated translation is not yet perfect):

```{r}
# English
forms_tables <- msmt_forms(what = "tables",
           language = "en")
forms_tables$M03$`02`
```

Function **msmt_forms** can be also used to search for forms and variables containing specific terms. Say, we search for variables containing "SEN".

```{r}
matching_forms <- msmt_forms(what = "search",
           language = "en",
           search_string = "SEN",
           exact_match = FALSE)
matching_forms
```

Finally, to facilitate quick browsing through forms, tables and variables, it is possible to use the "browser" option, returning a named list that can be easily browsed in R using the **$** operator to find variable codes. In the example bellow, we get variable names for the number of pupils with SEN in total and the number of girls with SEN.

```{r}
browser_forms <- msmt_forms(what = "browser", language = "en")

SEN_varname <- browser_forms$`Primary school report; form M03`$`Pupils with SVP total - number; etc. ; tab. 02`$`Pupils with SVP total - number; var. r02012`
SEN_girl_varname <- browser_forms$`Primary school report; form M03`$`Pupils with SVP total - number; etc. ; tab. 02`$`Pupils with SEN total - of which girls; var. r02013`

SEN_varname
SEN_girl_varname
```
## MSMT identifiers

Most **MSMT** data come with four identifiers, three of them nested:

* **year** The year represented in the data.
* **red_izo** The identifier of a headmaster or a director office, sometimes managing multiple schools.
* **izo** Nested under *red_izo*. The identifier of individual facility, e.g. under one *red_izo*, there can be an *izo* for a primary school and another izo for a preschool.
* **p_izo** Nested under *izo*. The location identifier, e.g. under one *izo* of a primary school, there can be two separate buildings with their own *p_izo*.

## Downloading MSMT data

Let's have a look at the gender of children with Special Educational Needs. We will first download the data for primary schools from form **M03**. Then, we can read the data, using function **msmt_read_data**.

```{r, message=FALSE, results = FALSE, warning=FALSE}
data_locations <- msmt_download_data(form = "M03", 
                                     years = 2016:2022) # We start with 2016, 
data_list <- msmt_read_data(data_locations = data_locations) # as many variables
                                              # are not available in 2015 forms.
```

The data in forms come in two formats, a wide format, where each row represent a unique school facility and long format, where multiple rows can represent the same facility. Most variables are reported in the wide format, long format usually contains counts of pupils per nationality and similar information. Here, we select the wide data and choose variable **year** and our pair of selected variables. It is possible to replace variable codes by using **msmt_names**.

```{r}
# Extract only wide data and 
data_use <- data_list$wide[,c("year", SEN_varname, SEN_girl_varname)]
# Assign labels
data_named <- msmt_names(data = data_use, form = "M03", language = "en")
names(data_named)
```

Now, we can start studying the gender structures of SEN in primary education.

```{r, message=FALSE, warning=FALSE}
library(tidyverse)

data_named %>%
  group_by(year) %>%
  summarise_all(sum, na.rm = TRUE) %>%
  pivot_longer(-year) %>%
  ggplot(aes(x = year, y = value, color = name)) +
  geom_path(aes(group = name)) +
  theme(legend.position = "bottom") +
  scale_y_continuous(labels = scales::comma) +
  theme()
```

## Spatial dimensions of MSMT data

Schools are situated not only in time, but also in space. **MSMT** contains several functions that facilitate the use of space in analysis. Let's have a look at the spatial dimensions of special educational needs. First, we download the address data and complete it with coordinates:

```{r, message=FALSE, results = FALSE, warning=FALSE}
addresses_prague <- msmt_addresses(NUTS3 = "CZ010") # Download address data
                                                    #(Prague only, to save time)
addresses_prague_coord <- msmt_coordinates(addresses_prague) # Append the data 
                                                             # with coordinates
addresses_prague_coord_elschool <- addresses_prague_coord %>% # Extract primary 
  filter(p_izo %in% data_list$wide$p_izo)                     # schools using  
                                  # the building identifier from our data on SEN
```

Having the coordinates, we can situate our schools in space.

```{r, echo = FALSE, message=FALSE, warning=FALSE}
library(RCzechia)

parts <- RCzechia::casti() %>%
  filter(NAZ_OBEC == "Praha")

data_coord <- addresses_prague_coord_elschool %>%
  select(p_izo, MistoRUAINKod_X, MistoRUAINKod_Y) %>%
  na.omit() %>%
  distinct() 

coord_object <- sf::st_as_sf(data_coord, 
                             coords = c("MistoRUAINKod_X", "MistoRUAINKod_Y"), 
                             crs = sf::st_crs(parts)) %>%
  sf::st_intersection(parts)

parts %>%
  ggplot(aes(geometry = geometry)) +
  geom_sf() +
  geom_sf(data = coord_object) +
  theme_void()
```

**MSMT** also contains a function **msmt_distances** that computes a matrix of distances between schools. The matrix can be used for several purposes. First, we can study the proximity of schools to each other itself, for instance, to simply see, how far every school is from it's fifth nearest school: 

```{r, message=FALSE, warning=FALSE}
library(stats)
msmt_dmat <- msmt_distances(id = data_coord$p_izo, # Get distance matric
                            X = data_coord$MistoRUAINKod_X, 
                            data_coord$MistoRUAINKod_Y)

clst_dmat <- as.dist(drop_units(msmt_dmat[coord_object$p_izo, # Select schools 
                                          coord_object$p_izo]))

min_dists <- clst_dmat %>% 
  apply(MARGIN = 1, function(x){sort(x)[5]}) # For every school, get distance to
                                             # fifth nearest school
```

```{r, echo = FALSE}
ggplot(tibble(x = min_dists), aes(x = x)) +
  geom_density(fill = "steelblue") +
  geom_rug() +
  labs(x = "distance")
```

Second, we can see, how school cluster in space. Given that the Czech school system is very fragmented with many extremely small schools, creating groups of schools for analytic purposes can be very helpful. Functions such as **hclust** or **kmeans** can be very helpful in this regard.

```{r}
cluster_object <- hclust(clst_dmat,
                         method = "average")

a_groups <- cutree(cluster_object, h = 2000) 
```

```{r, echo = FALSE}
a_groups <- a_groups %>%
  sample(1:length(.), length(.), replace = FALSE)[.] %>%
  tibble(gr = .,
         mc = coord_object$NAZEV) %>%
  group_by(mc) %>%
  mutate(gr = as.numeric(ordered(gr))) %>%
  ungroup() %>%
  mutate(mc = as.numeric(ordered(mc))) %>%
  mutate_all(as.character)

parts %>%
  ggplot(aes(geometry = geometry)) +
  geom_sf() +
  geom_sf(data = coord_object %>%
            bind_cols(a_groups), 
          aes(color = mc,
              fill = gr),
          pch = 21,
          cex = 1.75,
          stroke = 1,
          show.legend = FALSE) +
  theme_void()
```

Finally, we can use the distances to match schools to other schools in their surroundings and analyze variables in their local contexts. Consider the following:

Each school functions in an area with $N_t$ special needs children, out of which $N_g$ are girls. Within the area, the children are somehow distributed among schools.

When there is no gender bias, the proportion of SEN girls in a school $\frac{n_g}{n_t}$ should be approximately equal to the total proportion in the area $\frac{N_g}{N_t}$. 

Matching every school to it's neighboring schools allows us to use the original two variables and construct a gender bias in SEN index: $I = \frac{n_g}{n_t}-\frac{N_g}{N_t}$. Negative values mean, that the school has a lower proportion of girls, than it would have under a gender-independent distribution across schools. Positive values on the other hand mean, that the school has more girls, than it would have under a gender-neutral distribution.

The index can be completed with a $\chi^2$ statistical test of independence.

As shown in the plot bellow, there is only one primary school in Prague, displaying above chance levels of (at corrected $\alpha = .00005$) bias. This school over-represents SEN boys. In a closer inspection, this turns out to be a school specializing on children with behavioral disorders such as ADHD, more often diagnosed in boys.

```{r, message=FALSE, warning=FALSE}
msmt_nearest_schools <- msmt_nearest(msmt_dmat, 
                                     dist = 2000, # For every school, get all  
                                     format = "tibble")$dist # schools within
                                                             # a 2 km radius

values_t <- data_named %>% # Get number of SEN pupils for schools
  mutate(p_izo = data_list$wide$p_izo) %>%
  na.omit() %>%
  filter(year == "2022") %>%
  mutate(value = `Pupils with SVP total - number (M03; r02012)`)

values_g <- data_named %>% # Get numberf of SEN girls
  mutate(p_izo = data_list$wide$p_izo) %>%
  na.omit() %>%
  filter(year == "2022") %>%
  mutate(value = `Pupils with SEN total - of which girls (M03; r02013)`)

counts_t <- values_t$value %>% `names<-`(values_t$p_izo)
counts_g <- values_g$value %>% `names<-`(values_g$p_izo)

nearby_t <- msmt_nearest_schools[,-1] %>% # Compute total counts in radius
  apply(MARGIN = 1, FUN = function(x){sum(counts_t[x], na.rm = TRUE)})
nearby_g <- msmt_nearest_schools[,-1] %>%
  apply(MARGIN = 1, FUN = function(x){sum(counts_g[x], na.rm = TRUE)})

own_t <- counts_t[msmt_nearest_schools$ids] # Get counts for individual schools
own_g <- counts_g[msmt_nearest_schools$ids]

cross_tabs <- tibble(own_t, own_g, nearby_t, nearby_g) %>% # Get chisq p-values
  mutate(id = msmt_nearest_schools$ids) %>%
  na.omit() %>%
  pivot_longer(-id) %>%
  mutate(g1 = gsub(".+_", "", name),
         g2 = gsub("_.+", "", name)) %>%
  select(-name) %>%
  xtabs(value ~ g1 + g2 + id, data = .)

p_vals <- lapply(msmt_nearest_schools$ids,
                 function(x){
                   if(x %in% names(cross_tabs[1,1,])){
                     if(sum(cross_tabs[,,x]) != 0){
                       chisq.test(cross_tabs[,,x])$p.value
                     }else{NA}
                   }
                   else{NA}}) %>%
  unlist() %>%
  `names<-`(msmt_nearest_schools$ids)

tibble(index = (own_g/own_t)-((own_g + nearby_g)/(own_t + nearby_t)), # index
       p_val = p_vals) %>%
  arrange(index) %>%
  mutate(sig = p_val < .00005) %>% # corrected alpha level
  na.omit() %>%
  mutate(order = 1:n()) %>%
  ggplot(aes(x = order, y = index)) +
  geom_point(aes(color = sig, size = ifelse(sig, 1,.1)), show.legend = FALSE)
```
